---
title: "Using SimSpin"
author: "Katherine Harborne"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

``` {r, message=FALSE, warning=FALSE} 
library(SimSpin)
```

The purpose of this package is to measure the observable spin parameter $\lambda_R$ for simulated galaxy models with an effort to provide a method of communication between observers and computational astrophysicists. The aim is to address a simple question: is the true spin parameter, $\lambda$, and the one that we measure observationally, $\lambda_R$, comparable? 

This notebook will walk through a few of the basic steps that we can use to analyse a galaxy model. Here, we will use `filename = "~/SimSpin/inst/extdata/S0_vignette"` to explore the functions available in this package. This is the Gadget snapshot for a $10^{10} M_{\odot}$, S0 galaxy containing 50,000 disk and 75,000 bulge particles of equal mass. This is shown below. 

``` {r} 
galaxy_data = snapshot::snapread("~/LambdaR/inst/extdata/S0_vignette")
```

``` {r, echo=FALSE, fig.width=5, fig.height=5, fig.align="center"}
magicaxis::magplot(galaxy_data$part$x[(galaxy_data$head$Npart[3]+1):(galaxy_data$head$Npart[3]+galaxy_data$head$Npart[4])], galaxy_data$part$y[(galaxy_data$head$Npart[3]+1):(galaxy_data$head$Npart[3]+galaxy_data$head$Npart[4])], pch = '.', xlim = c(-20, 20), ylim = c(-20, 20), xlab = "x-position, kpc", ylab = "y-position, kpc", col = rgb(0, 0, 0, 0.05))
points(galaxy_data$part$x[1:galaxy_data$head$Npart[3]], galaxy_data$part$y[1:galaxy_data$head$Npart[3]], col = rgb(1, 0, 0, 0.05), pch = '.')
legend("topleft", inset=c(0.05,0.05),c("Bulge particles", "Disk particles"), col= c("black", "red"), pch= c(19, 19), bty="n", cex=0.9)

magicaxis::magplot(galaxy_data$part$x[(galaxy_data$head$Npart[3]+1):(galaxy_data$head$Npart[3]+galaxy_data$head$Npart[4])], galaxy_data$part$z[(galaxy_data$head$Npart[3]+1):(galaxy_data$head$Npart[3]+galaxy_data$head$Npart[4])], pch = '.', xlim = c(-20, 20), ylim = c(-20, 20), xlab = "x-position, kpc", ylab = "z-position, kpc", col = rgb(0, 0, 0, 0.05))
points(galaxy_data$part$x[1:galaxy_data$head$Npart[3]], galaxy_data$part$z[1:galaxy_data$head$Npart[3]], col = rgb(1, 0, 0, 0.05), pch = '.')
```

## Analysing the simulation

To begin, we can use the `sim_analysis()` function to calculate a few useful quantities for our model. Using this function, we can assess several spatial and kinematic properties such as:

* the mass distribution
* the `log(`density`)` distribution
* the circular velocity distribution
* the velocity anisotropy ($\beta$) distribution
* the Bullock spin parameter ($\lambda$) distribution

```{r}
sim_data = sim_analysis(filename = "~/SimSpin/inst/extdata/S0_vignette", 
                        bin_type = "r",
                        ptype    = NA,  # considering all particles
                        rmax     = 200, # maximum radius considered, kpc
                        rbin     = 200) # number of radial bins considered
```

When using `sim_analysis()`, we specify the snapshot `filename`, the type of particle we want to analyse `ptype = NA (all), 1 (gas), 2 (dark matter), 3 (disc), 4 (bulge), 5 (stars), 6 (boundaries)` or any present combination, `c(3,4)`, the maximum radius `rmax` in kpc we want to consider and the number of radial bins that we wish to divide the galaxy into, `rbin`. In the above example, all particles are grouped into bins of width `rmax / rbin = 1` kpc. Because we have specified `binddir = "r"` our bins are spherical shells. If we wished to study the distributions of certain components in cylindrical coordinates, we would specify either `bindir = "cr"` (cylindrical radial coordinated) or `="z"` (for 1D coordinates off the surface of the disc).

Below, a few of the possible distributions are plotted using the data produced by `sim_analysis()`.

``` {r, echo=FALSE, fig.width=3, fig.height=3}
magicaxis::magplot(sim_data$r, sim_data$Mass, type="l", lwd=3, col="red", xlab="r, kpc", ylab=expression("Mass, 10"^{10}*" M"['\u0298']))
magicaxis::magplot(sim_data$r, sim_data$logp, type="l", lwd=3, col="blue", xlab="r, kpc", ylab=expression("log("*paste(rho)*"),  10"^{10}*" M"['\u0298']*"kpc"^{-3}))
magicaxis::magplot(sim_data$r, sim_data$B, type="p", pch=20, col="orange", xlab="r, kpc", ylab=expression(paste(beta)), ylim=c(-4, 1))
magicaxis::magplot(sim_data$r, sim_data$lambda, type="l", lwd=3, col="green", xlab="r, kpc", ylab=expression(paste(lambda)))
```

Alternatively, you can consider single components of the model. For example, if we wished to examine the properties of the galaxy disk, we specify `ptype = 3`. The same could be done to examine the bulge by specifying `ptype = 4`. If you wish to examine a combination of parameters, this is simply specified by `ptype = c(2,3)` to give, for example, the dark matter and disc components - though this vignette galaxy model contains no dark matter so this will result in an error as shown. 

```{r, error=TRUE, purl=FALSE}
disk_data  = sim_analysis(filename = "~/LambdaR/inst/extdata/S0_vignette", 
                          ptype    = 3,      # considering just the disk particles
                          rmax     = 200, 
                          rbin     = 200)

bulge_data = sim_analysis(filename = "~/LambdaR/inst/extdata/S0_vignette", 
                          ptype    = 4,      # considering just the bulge particles
                          rmax     = 200, 
                          rbin     = 200)

dm_data    = sim_analysis(filename = "~/LambdaR/inst/extdata/S0_vignette", 
                          ptype    = c(2,3), # considering just the dark matter particles
                          rmax     = 200, 
                          rbin     = 200)
```
``` {r, echo=FALSE, fig.width=3, fig.height=3} 
magicaxis::magplot(disk_data$r, disk_data$Mass, type="l", lwd=3, col="red", xlab="r, kpc", ylab=expression("Mass, 10"^{10}*" M"['\u0298']), ylim=c(0,0.65))
legend("bottomright", inset=c(0.05,0.05),c("Disk"), bty="n", cex=0.9)

magicaxis::magplot(bulge_data$r, bulge_data$Mass, type="l", lwd=3, col="blue", xlab="r, kpc", ylab=expression("Mass, 10"^{10}*" M"['\u0298']), ylim=c(0,0.65))
legend("bottomright", inset=c(0.05,0.05),c("Bulge"), bty="n", cex=0.9)
```

## "Observing" the simulation

An important method of this package is to take mock integral field unit (IFU) spatial observations of galaxy models. The following sections will take you through how to create a series of spatial IFU images for your simulation. The functions attempt to be as general as possible, with options allowing the user to specify the shape of the CCDs within your chosen telescope, the field of view, the spatial and velocity pixel scale and pivot wavelength of the filter used during the "observation". Once the IFU data cube has been created, this can be analysed as an observer would real data and the spin parameter $\lambda_R$ and the observed inclination can be calculated. 

### Preparing the simulation data for observation

In order to arrange the data in the way that a telescope would collect it, we use the `obs_data_prep()` function. Below, we prepare the `S0_vignette` model for observation using the blue filter on SAMI of the galaxy inclined at $75^o$ at redshift $z = 0.1$. 

```{r} 
SAMI_obs = obs_data_prep(filename     = "~/LambdaR/inst/extdata/S0_vignette",
                         ptype        = NA,         # considering all particles
                         r200         = 200,        # virial radius, kpc
                         z            = 0.1,        # galaxy redshift
                         fov          = 15,         # telescope field of view, arcseconds
                         ap_shape     = "circular", # arrangement of the CCDs
                         pivot_wvl    = 4500,       # central filter wavelength, angstroms
                         pixel_sscale = 0.5,        # spatial pixel scale, arcseconds
                         pixel_vscale = 1.04,       # velocity pixel scale, angstroms 
                         inc_deg      = 75)         # projected inclination, degrees
```

Alternatively, we could observe our model using MaNGA or SAURON. 

```{r} 
MaNGA_obs   = obs_data_prep(filename  = "~/LambdaR/inst/extdata/S0_vignette",
                            ptype     = NA,  
                            r200      = 200,
                            z         = 0.1,         
                            fov       = 32,          
                            ap_shape  = "hexagonal", 
                            pivot_wvl = 5200,
                            pixel_sscale = 0.5,
                            pixel_vscale = 0.5, 
                            inc_deg   = 75)

SAURON_obs = obs_data_prep(filename     = "~/LambdaR/inst/extdata/S0_vignette",
                           ptype        = NA,   
                           r200         = 200,
                           z            = 0.1,         
                           fov          = 37,          
                           ap_shape     = "square", 
                           pivot_wvl    = 4500, 
                           pixel_sscale = 0.5,
                           pixel_vscale = 1.1, 
                           inc_deg      = 75)       
```

The output information from this function gives a `data.frame` that can be used to plot the spatial properties of the observed galaxy model; it contains all of the particle data from the original Gadget snapshot file along with the projected line of sight spatial (`$galaxy_obs$z_obs` and `$galaxy_obs$r_obs`) and velocity (`$galaxy_obs$vy_obs`) properties. These can be used to plot the particles that would be viewed by the telescope, as shown below. This function also returns the number and size of spatial bins in kpc and arcseconds (`$sbin`, `$sbinsize` and `$pixsize`), the velocity bins (`$vbinsize`) and the angular size of each pixel for easy conversion (`$angular_size`) as required for constructing the IFU cube.
